---
title: "Decision Trees"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Decision Trees

 For this example, I will use the Kaggle Titantic Test and Train datasets. Let's read these in and clean them first. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, results ='show',include=TRUE,messages=FALSE)




TitanicTrainData <- read.csv("Titanic_Training_Data.csv", na.string=c(""))
##print it...
(head(TitanicTrainData, n=5))
filename="Titanic_Testing_Data.csv"
## Note: na.string=c("") will replace empty with NA
TitanicTestData <- read.table(filename, sep=",", header=TRUE, na.string=c(""))
##print it...
(head(TitanicTestData,n=5))

```

# Data Cleaning

Recall that Decision Trees can only split nominally. This means that any quantitative (numerical) variables will have to be discretized/binned or removed; numerical ordinal data or data with too many categories should also be consolidated.

Let's consider each attribute. Passenger ID - not useful - so we can remove it. If we want to look up a passenger ID later, we can remove the column after we clean and prep everything, so that it stays aligned. For this case, I do not need it. So I will remove it now. I am also going to remove the Cabin, Name, and the Ticket. I like to use temp data frames so as to keep the originals.

```{r clean1} 


## Clean and prepare the data

## Look at the structure
(str(TitanicTestData))
(str(TitanicTrainData))


TempTitanicTrain <- TitanicTrainData
TempTitanicTest <- TitanicTestData

# Remove PassengerID, Name, Ticket, and Cabin 
TempTitanicTrain <-TempTitanicTrain[ , -which(names(TempTitanicTrain) %in% 
                             c("PassengerId","Name","Ticket","Cabin"))]
(head(TempTitanicTrain, n=10))

TempTitanicTest <- TempTitanicTest[ , -which(names(TempTitanicTest) %in% 
                             c("PassengerId","Name","Ticket","Cabin"))]
(head(TempTitanicTest, n=10))

```

Next - check how many NAs or missing values.


```{r isNA}

(head((is.na(TempTitanicTrain))))
(sum(is.na(TempTitanicTrain)))
(head((is.na(TempTitanicTest))))
(sum(is.na(TempTitanicTest)))

```

The AGE variable has a lot of missing items - let's  think about what do do about this. We can remove the age variable, we can remove the rows with NA, or we can try to fill in the missing ages with the age mean or median. Filling in the values is not always a good idea because
we are trying to build a predictor (a decision-maker). Using false ages may generate potentially incorrect results. So - we can remove the column or remove the rows with NA.
There is no perfect choice. So let us see how many rows we have left after removing the rows with NA.

```{r completeRows}

## How many rows are complete?
cat("The Titanic Test data has a total of ", nrow(TempTitanicTest), "rows.")
cat("The Titanic Train data has a total of ", nrow(TempTitanicTrain), "rows.")
TotalCompleteRowsTrain <- (nrow(TempTitanicTrain[complete.cases(TempTitanicTrain),]))
TotalCompleteRowsTest <- (nrow(TempTitanicTest[complete.cases(TempTitanicTest),]))
cat("The Titanic Train data has a total of ", TotalCompleteRowsTrain, "complete rows.")
cat("The Titanic Test data has a total of ", TotalCompleteRowsTest, "complete rows.")

## The above tells us that we will still have a large testing and training
## set - even if we remove all rows with NA

TempTitanicTrain <- TempTitanicTrain[complete.cases(TempTitanicTrain),]
TempTitanicTest <- TempTitanicTest[complete.cases(TempTitanicTest),]
## double check - both of these should be 0 - which they are
(nrow(TempTitanicTrain[!complete.cases(TempTitanicTrain),]))
(nrow(TempTitanicTest[!complete.cases(TempTitanicTest),]))

```

## Now its time to discretize/ bin some of the variables where it may be appropriate. 

```{r bin}
## Let's look at the str and tables
(str(TempTitanicTest))
(str(TempTitanicTrain))
## change "survived" to a factor
TempTitanicTrain$Survived=factor(TempTitanicTrain$Survived)
##Pclass is the classification of the ticket - 1st, 2nd, 3rd.
## Make this into a factor
TempTitanicTrain$Pclass=factor(TempTitanicTrain$Pclass)
TempTitanicTest$Pclass=factor(TempTitanicTest$Pclass)
## Age is quantitative and must be binned and discretized
## Check Age for errors and look at values
(freq=table(TempTitanicTrain$Age))
## We see that there are some incorrect ages: .42, .67, .75, .83, and .92 are not correct. 
## Remove those rows first
(freq=table(TempTitanicTest$Age))  ## Same errors in Test data

## Place NA for any ages that are < 1
TempTitanicTrain$Age <- ifelse(TempTitanicTrain$Age < 1, "NA", TempTitanicTrain$Age)
(freq=table(TempTitanicTrain$Age))
TempTitanicTest$Age <- ifelse(TempTitanicTest$Age < 1, "NA", TempTitanicTest$Age)
(freq=table(TempTitanicTest$Age))
## Remove NAs

TempTitanicTrain <- TempTitanicTrain[complete.cases(TempTitanicTrain),]
TempTitanicTest <- TempTitanicTest[complete.cases(TempTitanicTest),]

## Now we can discretize the Age
TempTitanicTrain$Age[TempTitanicTrain$Age <= 22] <- 1
TempTitanicTrain$Age[TempTitanicTrain$Age > 22 & TempTitanicTrain$Age <=38] <- 2
TempTitanicTrain$Age[TempTitanicTrain$Age > 38] <- 3
TempTitanicTrain$Age=factor(TempTitanicTrain$Age)
(TempTitanicTrain$Age)

TempTitanicTest$Age[TempTitanicTest$Age <= 22] <- 1
TempTitanicTest$Age[TempTitanicTest$Age > 22 & TempTitanicTest$Age <=38] <- 2
TempTitanicTest$Age[TempTitanicTest$Age > 38] <- 3
TempTitanicTest$Age=factor(TempTitanicTest$Age)
(TempTitanicTest$Age)

## check it
(freq=table(TempTitanicTrain$Age))
(freq=table(TempTitanicTest$Age))

##Look at the str again
(str(TempTitanicTest))
(str(TempTitanicTrain))

## Now we will discretize Sibsp (number of siblings on board)
## and Parch (number of parents or children)
## WHile I will not do this here - it might also be interesting to 
## add these to create a vew attribute called Family

(freq=table(TempTitanicTrain$SibSp))
(freq=table(TempTitanicTest$SibSp))

(freq=table(TempTitanicTrain$Parch))
(freq=table(TempTitanicTest$Parch))

## Given that so much of the data is at 0 or 1, I will group
## SibSp and Parch so that they are 0 (none) or 1 (one or more)

TempTitanicTrain$SibSp[TempTitanicTrain$SibSp == 0] <- 0
TempTitanicTrain$SibSp[TempTitanicTrain$SibSp > 0] <- 1
TempTitanicTrain$SibSp=factor(TempTitanicTrain$SibSp)
(TempTitanicTrain$SibSp)

TempTitanicTrain$Parch[TempTitanicTrain$Parch == 0] <- 0
TempTitanicTrain$Parch[TempTitanicTrain$Parch > 0] <- 1
TempTitanicTrain$Parch=factor(TempTitanicTrain$Parch)
(TempTitanicTrain$Parch)

TempTitanicTest$SibSp[TempTitanicTest$SibSp == 0] <- 0
TempTitanicTest$SibSp[TempTitanicTest$SibSp > 0] <- 1
TempTitanicTest$SibSp=factor(TempTitanicTest$SibSp)
(TempTitanicTest$SibSp)

TempTitanicTest$Parch[TempTitanicTest$Parch == 0] <- 0
TempTitanicTest$Parch[TempTitanicTest$Parch > 0] <- 1
TempTitanicTest$Parch=factor(TempTitanicTest$Parch)
(TempTitanicTest$Parch)


(freq=table(TempTitanicTrain$SibSp))
(freq=table(TempTitanicTest$SibSp))

(freq=table(TempTitanicTrain$Parch))
(freq=table(TempTitanicTest$Parch))

## Continue the process...look at the currect str
(str(TempTitanicTest))
(str(TempTitanicTrain))

## The last step is to discretize the Fare
(freq=table(TempTitanicTrain$Fare))
(freq=table(TempTitanicTest$Fare))
## We can see that the range is within 0 - 550
## We could also have gotten the min and max

## Place into three groups: 1: Low, 2: Mediun, 3: High  Fare rate
TempTitanicTrain$Fare[TempTitanicTrain$Fare <= 10] <- 1
TempTitanicTrain$Fare[TempTitanicTrain$Fare > 10 & TempTitanicTrain$Fare <=30] <- 2
TempTitanicTrain$Fare[TempTitanicTrain$Fare > 30] <- 3
TempTitanicTrain$Fare=factor(TempTitanicTrain$Fare)
#(TempTitanicTrain$Fare)

TempTitanicTest$Fare[TempTitanicTest$Fare <= 10] <- 1
TempTitanicTest$Fare[TempTitanicTest$Fare > 10 & TempTitanicTest$Fare <=30] <- 2
TempTitanicTest$Fare[TempTitanicTest$Fare > 30] <- 3
TempTitanicTest$Fare=factor(TempTitanicTest$Fare)
#(TempTitanicTest$Fare)

(freq=table(TempTitanicTrain$Fare))
(freq=table(TempTitanicTest$Fare))
(str(TempTitanicTest))
(str(TempTitanicTrain))

## Our datasets are finally clean and ready to use.
## Let's move the temp to new dataframe names
CleanTest <- TempTitanicTest
CleanTrain <- TempTitanicTrain
```

# Decision Trees
The goal is to create decision tree from the training data that can classify a row as survive or not survive.Then, we will use the testing data to see how well the Decision Tree works. 

```{r decisionTree}

######################################### BUILD Decision Trees ----------------------------
#install.packages("rpart")
#install.packages('rattle')
#install.packages('rpart.plot')
#install.packages('RColorBrewer')
#install.packages("Cairo")
#install.packages("CORElearn")
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(Cairo)


# rpart will train the decision tree based on the train data. 
fit <- rpart(CleanTrain$Survived ~ ., data = CleanTrain, method="class")
summary(fit)


# Once trained we can test our tree on the test data.
predicted= predict(fit,CleanTest, type="class")
(head(predicted,n=10))
(head(CleanTest, n=10))
```


In the next segment we will illustrate the decision tree.

```{r display}

fancyRpartPlot(fit)
submit <- data.frame(PassengerGender = CleanTest$Sex, Survived = predicted)
(head(submit, n=10))
write.csv(submit, file = "TitanicPrediction.csv", row.names = FALSE)

## Let's reduce the tree size
fit2 <- rpart(Survived ~ Pclass + Sex + Age,
             data=CleanTrain,
             method="class", 
             control=rpart.control(minsplit=2, cp=0))

fancyRpartPlot(fit2)

## Save the Decision Tree as a jpg image
jpeg("DecisionTree_Titanic.jpg")
fancyRpartPlot(fit2)
dev.off()

```

